\emph{icepack}: a new glacier flow modeling package in Python, version 1.0

### 2021-03-16 reviews/response-1.tex

\documentclass{article}

\usepackage{amsmath}
%\usepackage{amsfonts}
\usepackage{amsthm}
%\usepackage{amssymb}
%\usepackage{mathrsfs}
%\usepackage{fullpage}
%\usepackage{mathptmx}
%\usepackage[varg]{txfonts}
\usepackage{color}
\usepackage[charter]{mathdesign}
\usepackage[pdftex]{graphicx}
%\usepackage{float}
%\usepackage{hyperref}
%\usepackage[modulo, displaymath, mathlines]{lineno}
%\usepackage{setspace}
%\usepackage[titletoc,toc,title]{appendix}
\usepackage{natbib}

%\linenumbers
%\doublespacing

\theoremstyle{definition}
\newtheorem*{defn}{Definition}
\newtheorem*{exm}{Example}

\theoremstyle{plain}
\newtheorem*{thm}{Theorem}
\newtheorem*{lem}{Lemma}
\newtheorem*{prop}{Proposition}
\newtheorem*{cor}{Corollary}

\newcommand{\argmin}{\text{argmin}}
\newcommand{\ud}{\hspace{2pt}\mathrm{d}}
\newcommand{\bs}{\boldsymbol}
\newcommand{\PP}{\mathsf{P}}
\let\divsymb=\div % rename builtin command \div to \divsymb
\renewcommand{\div}[1]{\operatorname{div} #1} % for divergence
\newcommand{\Id}[1]{\operatorname{Id} #1}

\begin{document}

\title{Response to review \#1}
\author{}
\date{}

\maketitle

Thank you Doug for this helpful review!
Original comments are in blue, our responses in black.

\section*{Summary}

\textcolor{blue}{In ‘icepack: a new glacier flow modeling package in Python, version 1.0’, Shapero and co-authors present a promising new ice sheet modeling framework. The
framework contains mechanisms for solving both the prognostic mass balance
equations for updating ice sheet geometry, as well as diagnostic solvers for approximations to the non-linear Stokes’ equations. Throughout both the software
and the manuscript describing it, the authors focus on ensuring usability (readability), a trait that is bound to make this software (and paper) frequently used.
Despite its accessibility, the capabilities of icepack are already impressive, made
all the more so by its explicit design prioritization of easy extensibility.
I find the manuscript to be exceptionally well-crafted, and I think that it
could be published as is. That said, I offer a few suggestions, comments, and
points of clarification below.}

\section*{Minor points}

\begin{description}
\item[L15-18] \textcolor{blue}{It would be nice to have a cited example or three for each of these
suggested use cases. This would help the reader identify the kinds of
practical problems where icepack might fill a need.}

\item[L74] \textcolor{blue}{A low aspect ratio isn’t really an approximation; it’s an existential fact.
The approximation that the first order approximation makes is that vertical resistive stresses (or bridging stresses as they are referred to later
in this manuscript) are small, pressure is hydrostatic, and bed slopes are
small.}

\item[L92] \textcolor{blue}{It would be useful to offer a reference regarding an anisotropic fluidity.}

\item[L100] \textcolor{blue}{Not clear where the Legendre transform enters: the viscous and frictional
dissipation can be read off from Eq. 4.
Eqs. 6 and 12 While I understand that it is convenient to manipulate the
action to reflect the algebraic manipulations to yield the analytical SIA
solution, the break in symmetry between Eq. 6 and Eq. 12 is frustrating,
1given that they both are name ‘gravity’, and that they should in some
sense be the same regardless of which strain rates are assumed to be zero.}

\item[L144] \textcolor{blue}{That the terminal potential term, Eq. 13 disappears is not obvious. This
should probably be clarified, since many readers will be surprised by this.}

\item[L152] \textcolor{blue}{Cite the method of manufactured solutions.}

\item[L159] \textcolor{blue}{The benefit to avoiding complicated 3D meshing should not be under-
stated, in addition to the reduction in the cost of computational solution.}

\item[L188] \textcolor{blue}{This is a bit of a red herring, given that impenetrability is a natural
boundary condition on the incompressibility equation. It can go right
in the action principle, no extra Lagrange multiplier (besides pressure)
needed.}

\item[L367] \textcolor{blue}{I’m surprised by this. For challenging geometries, I’ve always needed to
stabilize even when using implicit Euler. Are you sure that implicit Euler
is unconditionally stable even for non-linear advection like this?}

\item[L453] \textcolor{blue}{Many advances have occurred in the last 5 years regarding gradient de-
scent due to its necessity for optimizing neural networks. These may yet
be useful in this context if you have to deal with a large scale optimization
problem where forming the Hessian becomes prohibitive.}
\item[L459] \textcolor{blue}{Gauss-Newton needs a reference if you’re not going to describe it here.
Eq. 34 It’s worth noting that the Schoof law is phenomenological, and was
selected because it has the right shape and obeys Iken’s bound. As such,
if your sliding law obeys Iken’s bound and looks right, then it’s not any
less valid.}
\item[L580] \textcolor{blue}{I think that having to specify a variational principle for the sliding law
is useful because it guarantees a law that is positive (semi-)definite, as it
must be to be physical.}
\item[L701–713] \textcolor{blue}{I enjoyed reading this paragraph, but I wonder if stabilizing the
Stokes’ equations is the best illustration of the point, given that icepack
does not in fact have Stokes’ equations implemented (although I imagine
it could be done in short order). Maybe stabilizing a transport equation
would be a more appropriate case?}
\end{description}

\end{document}

### 2021-03-16 reviews/response-2.tex

\documentclass{article}

\usepackage{amsmath}
%\usepackage{amsfonts}
\usepackage{amsthm}
%\usepackage{amssymb}
%\usepackage{mathrsfs}
%\usepackage{fullpage}
%\usepackage{mathptmx}
%\usepackage[varg]{txfonts}
\usepackage{color}
\usepackage[charter]{mathdesign}
\usepackage[pdftex]{graphicx}
%\usepackage{float}
%\usepackage{hyperref}
%\usepackage[modulo, displaymath, mathlines]{lineno}
%\usepackage{setspace}
%\usepackage[titletoc,toc,title]{appendix}
\usepackage{natbib}

%\linenumbers
%\doublespacing

\theoremstyle{definition}
\newtheorem*{defn}{Definition}
\newtheorem*{exm}{Example}

\theoremstyle{plain}
\newtheorem*{thm}{Theorem}
\newtheorem*{lem}{Lemma}
\newtheorem*{prop}{Proposition}
\newtheorem*{cor}{Corollary}

\newcommand{\argmin}{\text{argmin}}
\newcommand{\ud}{\hspace{2pt}\mathrm{d}}
\newcommand{\bs}{\boldsymbol}
\newcommand{\PP}{\mathsf{P}}
\let\divsymb=\div % rename builtin command \div to \divsymb
\renewcommand{\div}[1]{\operatorname{div} #1} % for divergence
\newcommand{\Id}[1]{\operatorname{Id} #1}

\begin{document}

\title{Response to review \#2}
\author{}
\date{}

\maketitle

\section*{Summary}

\textcolor{blue}{In this manuscript, the authors introduce a new land-ice modeling software package known as icepack.
icepack is written in Python on top of the Firedrake library, which uses the domain-specific Unified Form
Language (UFL) and provides a high-level symbolic description of the problem to facilitate the add of new
physics and/or equations. The intended user base of icepack is the glaciological community, in particular,
glaciologists who may not have an extensive background/training in computational science. The idea is to
provide a code that would be easy to use/develop by this class of prospective users. Following a description
of the code, some numerical examples are presented to demonstrate the method’s capabilities and accuracy.
The manuscript in question is well-written and interesting to read. Addressing usability is noteworthy
and something that not enough authors in glaciology/climate science address. I personally have some
qualms with some of the philosophy described by the authors, namely I worry about folks who are not
familiar with numerical methods developing an application code in which a lot of what is under the
code is hidden from them in some sense. In an ideal world, one would have glaciologists working with
computational scientists to help them pick the right solvers, discretizations, etc., for their problem. The
authors are correct that some solver options are for optimizing performance, which is secondary to getting
the code/model running; but there are also solver/algorithm choices that depend very much on the physics
(e.g., CG is only valid for symmetric problems) - is icepack designed so as to prevent the naive user from
inadvertently using the incorrect default setting for their problem? I will assume it is to the extent it can
be, and that the authors’ argument is that the default hidden settings are likely to do less damage than
some arbitrary settings a user might put in his/her input file without knowing what they are doing. Also,
I realize that the reality is that many glaciologists do not have strong ties to computational scientists, and
still wish to make progress in numerical modeling of land-ice; therefore, I will not focus too much on much
“philosophical” perspective described above.
Another qualm I have about the paper has to do with performance - I am skeptical whether icepack
can really be performant if one tries to run it on continental scale problems, and it is not clear to me if
the code is even parallel. I think the intention may be to use icepack as more of a sandbox for prototyping
small problems (similar to FeniCS), in which case, this is not a huge deal.
Despite the above concerns, I like this paper and see it published in GMD. I like in particular the
idea of describing all the equations using the variational principle/action functional and having everything
else propagate from there - not enough people do this. I do, however, feel that there is a lot of missing
information in various parts of the paper, which should be filled in in the revision prior to the paper being
suitable for publication. Please see below my enumerated list of questions/comments to address in the
revision.}

\section*{Specific comments}

\begin{enumerate}
\item \textcolor{blue}{The authors suggest that C++ makes it inherently difficult to add new physics/PDEs (e.g., on p. 10
1and p. 20), which I somewhat disagree with. One advancement that can make a C++-based code
easy to add to is Automatic Differentiation (AD) - with AD, one can effectively code the weak form
of the residual within a C++ code and AD will handle the rest, making it very easy to add new
physics. An example is the Albany/Land Ice model (previously known as Albany/FELIX) of Tezaur
et al. (https://doi.org/10.5194/gmd-8-1197-2015). I think it would be worth mentioning that
there have been efforts like Albany/Land Ice out there to make C++-based codes more accessible to
users of varied backgrounds. I agree that even an “easy-to-use” C++ code will be more difficult and
more intimidating than a Python code, so I am not trying to minimize the authors’ efforts at all.}
\item \textcolor{blue}{Can the authors comment on the overhead of the symbolic descriptions/manipulations done by their
framework? This sounds potentially like it would be very expensive. How does the cost compare to
automatic differentiation, for example? A broader question is: is computational performance/cost a
concern for users of icepack, or is it intended to be a “sandbox” in which performance is secondary
to being able to code up something “quick-and-dirty” for initial prototyping?}
\item \textcolor{blue}{I was a little bit confused about the reference of the FO Stokes-based model in this paper as a “hybrid
model”. I see that it is hybrid in the sense that you have a different discretization in the horizontal and
vertical direction, but there are also hybrid ice models that use different PDEs in different domains,
e.g., the ISCAL model of Ahlkrona et al. (https://doi.org/10.1016/j.quascirev.2016.01.032).
Is the term “hybrid model” a common name for the approach in Section 2.2.3? Perhaps it is and I
am not aware of it.}
\item \textcolor{blue}{Does the hybrid model described in the paper have the same applicability as say the First Order
Stokes model? Can it be used for both Greenland and Antarctica at continental scales?}
\item \textcolor{blue}{Section 2.2.1: in my opinion, the authors do not provide sufficient justification for the penalty term,
equation (7). They describe this as something that is added to smooth over artifacts - this would be
needed based on the discretization, which there is little discussion of. The authors should state what
order finite elements they are using - I presume they are linear, and that this is why the stabilization
is needed? Why is stabilization needed only for the SIA? I think these things should be made clear.}
\item \textcolor{blue}{Section 2.2.2: there is some imprecision here in equation (13) - you have not defined anywhere that $\Gamma$
is the boundary, and which boundary you are referring to. One can figure it out, but it is not precise.
$\Omega$ is not defined either though one will assume invariably that this is an open bounded domain in 2D
or 3D depending on which approximation one is looking at.}
\item \textcolor{blue}{The boundary conditions are not discussed very rigorously systematically - the authors seem to
sprinkle in some boundary conditions here and there. I think the boundary conditions need to be
given for each of the models at the time the models are presented - boundary conditions are needed
to complete the definition of each models.}
\item \textcolor{blue}{Certain terms in the equations I do not believe are defined anywhere, for instance, in equation (10),
there is no expression given for the strains $\dot\epsilon(u)$.
This is one of the things I cam across that need to
be made more precise.}
\item \textcolor{blue}{Section 2.2.3: the authors comment that higher degree polynomials can be used in the vertical layer
in the hybrid approach. What order is typically used?}
\item \textcolor{blue}{Section 2.2.3: this might be a naive question, but does Glen’s flow law come into the hybrid model?
I was expecting to see it there, but maybe I’m missing something.}
\item \textcolor{blue}{P. 10: the discussion here about substituting model components suggests it may be possible to use
different models in different regions and couple them (a la the ISCAL method). Is this possible, or
something that the authors are thinking to add to their model/code?}
\item \textcolor{blue}{Section 2.4: This section is very incomplete. You need to give the enthalpy equation and given the
Glen’s law expression as well, since it is mentioned.}
\item \textcolor{blue}{In my opinion, there is not enough discussion of the thickness equation (Section 2.1) and how it
is discretized. In typical ice sheet models, this equation is used to change the ice extent - one
meshes up a region of “potential” ice, and then uses the thickness to dynamically determine a mask
for ice-covered regions. Do you do something like this in your model? It should be discussed for
completeness. I think you maybe start to do this in Section 4.1, but it is very confusing and hard to
make the connection.}
\item \textcolor{blue}{What sort of meshes do you use in your model (in the horizontal dimension, for the hybrid one)?
Structured/unstructured? Hex/tet (quad/tri)?}
\item \textcolor{blue}{Section 3: It is not clear from the description what the inverse problem you are describing is for.
Is it to obtain parameters in the model like the basal friction using observational data of e.g. sur-
face velocity? There really needs to be more discussion here, and I personally would like to see
a mathematical statement of a representative inverse problem you are solving. It would be worth
citing the work Perego et al. on optimization-based inversion, if what you are doing is similar:
https://doi.org/10.1002/2014JF003181. BTW, the basal friction has not been defined, yet it is
discussed - it needs to be defined earlier, when talking about boundary conditions (which needs to
be added).}
\item \textcolor{blue}{Section 4.1: I find this section confusing. I assume you are talking about discretizing the thickness
equation here - that should be made clear. I disagree with several statements in this section as
well. “The simplest explicit timestepping schemes are unstable with CG finite elements” - if you are
talking about CFL stability, this is not true. You need to satisfy a CFL condition which could give
rise to very small time-steps but you can get the scheme to be stable. I’m also confused about the
notion of SUPG as a time-stepping scheme - I think of SUPG as a finite element approach to deal
with advection-dominated flow problems, for example, that does not have anything to do with time-
stepping. Maybe you are referring to upwinding? In any case, I think SUPG has nothing to do with
forward Euler, so the discussion about forward Euler requiring parameters is erroneous. Additionally,
I don’t understand the comment about implicit Euler smoothing out sharp discontinuities... I believe
explicit and implicit Euler have effectively the same diffusion and dispersion properties, so there
should not really be a difference between the schemes. Did the authors verify their time-stepper on
a manufactured problem to ensure that it was implemented correctly?}
\item \textcolor{blue}{Section 4.2: there is an approach discussed in Tezaur et al. (https://doi.org/10.5194/gmd-8-1197-2015)
for dealing with bad initial conditions in a Newton solver that relies on homotopy continuation that
would be worth citing. It is an alternate to the approach you describe that lets you get away with not
doing a line search for Newton. By the way, it should be no surprise that Newton is not converging
3without a line search - in general Newton is not guaranteed to converge from an arbitrary initial
guess without the line search.}
\item \textcolor{blue}{Section 4.3, lines 411-412: there are actually ways to construct weighted norms to deal with the issue
of DOFs having different orders of magnitude for the purpose of convergence.}
\item \textcolor{blue}{Section 4.4: Again, it is not clear to me what is your inverse problem. You need to state this explicitly
so it is clear.}
\item \textcolor{blue}{Section 4.5.1: you talk about problems defined on “extruded geometries” - do you ever use non-
extruded geometries? It has been shown in various references that there can be numerical problems
for land-ice solvers that do not use extruded geometries, e.g. Tezaur et al.}
\item \textcolor{blue}{Section 4.5.2: I think this section needs to be made earlier, and other BCs need to be added to that
discussion.}
\item \textcolor{blue}{Section 4.6: The authors mention running their code on 1 core. Is the code parallel - can it be run
on multiple cores? Are there any hope for performance portability of the code to take advantage of
emerging HPC architectures, e.g., GPUs?}
\item \textcolor{blue}{Section 4.6: Can you please clarify what you mean by the following statement? “Large problems,
such as continental-scale modeling, will require more sophisticated and possibly problem-specific
approaches”. I’m wondering in particular about the problem-specific approach part. There are
models like first order Stokes that can be used at the continental scale and they are not really
problem specific.}
\item \textcolor{blue}{p. 20: I don’t understand why you need to create an analytical expression of (32) using special
functions. Is this something specific to your framework, which requires expressions in a certain form
for the symbolic representation?}
\item \textcolor{blue}{It’s great that you have executable documentation in something easy-to-use such as Jupyter note-
books! (no need to address this comment)}
\item \textcolor{blue}{p. 29: are you considering putting in the first-order Stokes//Blatter Pattyn model into your code
framework?}
\item \textcolor{blue}{The methods do not discuss their code development/testing stance on icepack. How do users con-
tribute to the code - through pull requests? Is there regression/performance testing? Continuous
integration testing? These are all really important, especially if you have non-experts contributing
to the code!}
\end{enumerate}

\section*{Minor comments}

\begin{itemize}
\item \textcolor{blue}{“UFL” is not defined in the abstract.}
\item \textcolor{blue}{p. 4, line 90: A is also called the “flow factor”. I would mention here that it is usually a function of
the temperature, which comes from a different equation.}
\item \textcolor{blue}{p. 5, line 125: should be “checking”, not “check”.}
\item \textcolor{blue}{p. 5, line 126: I think it should be “Bueler” not “Beuler”, if I’m thinking of the right person.}
\item \textcolor{blue}{p. 5, line 147: change “we verified the correctness of the ice shelf model” to “we verified the correctness
of our implementation of the ice shelf model”.}
\item \textcolor{blue}{p. 10, line 270: change “we’ll” to “we will”.}
\end{itemize}

\end{document}

### 2021-03-17 icepack.tex
Most of reviewer 1 corrections

The most complex and most physically accurate model that we implement is the \emph{first-order} or \emph{Blatter-Pattyn} approximation, a 3D system describing the horizontal velocity \citep{blatter1995velocity, pattyn2003new}.
(In icepack we refer to our implementation of this equation as a \emph{hybrid model} as it includes both shear and plug flow modes, described below.)
The only approximation that the first-order model makes is that the flow has a low aspect ratio -- the thickness of the glacier is much less than its horizontal extent.
This approximation may be questionable around, say, the main trunk of Jakobshavn Isbrae in Greenland, which flows through a very deep and narrow trough.
Even Jakobshavn has an aspect ratio on the order of 1/5 and almost all glacier flows have an aspect ratio less than 1/10 or even 1/20.

@@

All the diagnostic models inherit a fundamental nonlinearity in the mechanics of ice flow.
For a Newtonian viscous fluid, the stress tensor $\tau$ and the strain rate tensor $\dot\varepsilon$ are linearly proportional to each other.
Glaciers, however, have a nonlinear constitutive relation:
\begin{equation}
    \dot\varepsilon = A|\tau|^{n - 1}\tau
\end{equation}
where $A$ is the \emph{fluidity} and $n$ is the \emph{Glen flow law exponent} \citep{cuffey2010physics}.
The most commonly-used value of the flow law exponent is $n \approx 3$ as determined from laboratory experiments.
Strictly speaking, the fluidity should be a tensor field, but is almost universally treated as a scalar.

All of the diagnostic models in icepack are described through \emph{variational} or \emph{action} principles \citep{dukowicz2010consistent}.
Rather than describe the velocity as the weak solution of a nonlinear PDE, an action principle instead states that the velocity minimizes a functional called the action.
The action consists of four terms:
\begin{align}
    & \text{action} = \iint\text{stress} \times \text{strain rate}\,dz\,dx - \int\text{basal friction} \times \text{sliding velocity}\,dx \nonumber \\
    & \quad - \iint\text{surface slope}\times\text{velocity}\,dz\,dx - \iint\text{ocean pressure}\times\text{velocity}\,dz\,d\gamma
    \label{action-functional}
\end{align}
where $dz\,dx$ denotes integration over the entire glacier, $dx$ denotes integration over the glacier footprint, and $dz\,d\gamma$ over the side wall boundary.
The action has units of power (energy/time) and can be related to the rate of decrease of the thermodynamic free energy.
Moreover, the energy lost to viscous and frictional heating can be calculated from the action and its Legendre transform.
The action principle can be viewed as a consequence of the Onsager reciprocity relations for systems near to equilibrium \citep{de2013non}.

@@

The shallow stream model can be derived from the Blatter-Pattyn approximation by assuming nearly plug flow, i.e. $\partial u/\partial z \ll \partial u/\partial x$.
The momentum equations can, again, be vertically integrated to obtain a 2D system of equations.
The class in icepack that represents this physics model is called \texttt{IceStream}.
The terms in the action functional are:
\begin{align}
    \text{viscosity} & = \frac{n}{n + 1}\int_\Omega hA^{-\frac{1}{n}}|\dot\varepsilon(u)|^{\frac{1}{n} + 1}dx \\
    \text{friction} & = \frac{m}{m + 1}\int_\Omega C|u|^{\frac{1}{m} + 1}dx \\
    \text{gravity} & = -\int_\Omega\rho_I gh\nabla s\cdot u\,dx \\
    \text{terminus} & = \frac{1}{2}\int_\Gamma(\rho_I gh^2 - \rho_Wgd^2)u\cdot \nu\,d\gamma
\end{align}
and the action functional is
\begin{equation}
    J = \text{viscosity} + \text{friction} - \text{gravity} - \text{terminus}.
\end{equation}
When the ice is floating, there are two simplifications: the friction coefficient $C$ is 0, and the surface elevation $s$ can be written in terms of the thickness $h$ as
\begin{equation}
    s = (1 - \rho_I / \rho_W)h.
\end{equation}
Additionally, the terminal stress term of the action disappears by applying integration by parts to the gravity term.
Since the action functional for ice shelves has fewer terms than for grounded ice streams, we have defined a separate \texttt{IceShelf} model class.
The ice shelf and ice stream models share common components, i.e. the viscosity and side wall stress.

We verified the correctness of the ice shelf model by checking that numerical results converge at the expected rate to an analytical solution for the velocity with a linearly sloping thickness in a rectangular domain \citep{greve2009dynamics}.
When there is basal friction, the shallow stream equations do not have a simple analytical solution.
To get an exact solution, we chose the ice velocity and thickness a priori and used the computer algebra system SymPy \citep{sympy} to generate a friction coefficient that makes these fields an exact solution.
We then checked that numerical results converge to this manufactured solution at the expected rate.

@@

Finally, the Gauss-Newton solver defines an approximation to the ``first-order'' part of the objective functional Hessian.
Each iteration of Gauss-Newton is more expensive than that of BFGS or gradient descent because it requires the solution of a more complex linear system than just the mass matrix.
The Gauss-Newton method converges fastest by far in virtually every test case we have found, in some instances by up to factor of 50.

@@

The original intercomparison project specified that participants could use the Weertman sliding law (equation \eqref{eq:weertman-sliding}) as well as two other sliding laws that transition to a more plastic rheology at high sliding speeds.
The first alternative sliding law consists of Weertman sliding until the stress reaches a critical value, at which point the constitutive relation transitions to perfect plasticity.
The second alternative is the Schoof sliding law \citep{schoof2005effect}:
\begin{equation}
    \tau_b = -\frac{C|u|^{1/m}\cdot \tau_c}{(C^m|u| + \tau_c^m)^{1/m}}\frac{u}{|u|},
    \label{eq:schoof-sliding}
\end{equation}
where the critical stress $\tau_c$ is a certain specified fraction of the water pressure in the subglacial hydrological system.
Sliding laws in icepack are not expressed directly, but rather as the the derivative of an action functional.
To implement the Schoof sliding law we need to know the antiderivative of equation \eqref{eq:schoof-sliding}.
Using a computer algebra system, we found that the antiderivative of this expression in terms of hypergeometric functions.
The Unified Form Language has several transcendental functions (sine, cosine, exponential, etc.), but it does not currently support hypergeometric functions.
Instead, we implemented a sliding relation that exhibits the important features of the Schoof law, i.e. it behaves like an $m = 3$ power law at low sliding speeds and $m = \infty$ at high sliding speeds, but which is more tractable algebraically.
Knowing the critical stress $\tau_c$ and the friction coefficient $C$, which has units of stress $\times$ speed${}^{-1/m}$, we can define a \emph{critical speed} $u_c$ as
\begin{equation}
    u_c = C^{-m}\tau_c^m.
\end{equation}
The power dissipation density for the Schoof-type sliding law that we use is
\begin{equation}
    P = \tau_c\left\{\left(u_c^{\frac{1}{m} + 1} + |u|^{\frac{1}{m} + 1}\right)^{\frac{m}{m + 1}} - u_c\right\}
    \label{eq:power-dissipation-modified-schoof-sliding}
\end{equation}
Figure \ref{fig:sliding-laws} shows a comparison of the original Schoof sliding law and the sliding law that arises as the derivative of the functional in equation \eqref{eq:power-dissipation-modified-schoof-sliding}.
The two have the same asymptotic behavior when the speed is much smaller or much larger than the critical speed; they differ in a relatively small range around the critical speed.
The relative difference in basal shear stress between the two sliding laws is less than 10\% throughout the entire range.
Although figure \ref{fig:sliding-laws} shows a comparison of both sliding laws with the same value of the critical speed, by using different values, the agreement between our sliding law can be brought into much closer agreement with the Schoof law.

@@

Using variational principles to express all constitutive laws is less flexible than specifying the sliding law directly and this is a distinct disadvantage.
We were nonetheless able to implement a sliding law that exhibits the important characteristics of the Schoof sliding law.
\citet{asay2016experimental} also suggest using a sliding law that transitions sharply to exact plasticity above the critical speed.
Expressing this sliding law in UFL requires a conditional in the velocity and is thus no longer differentiable, causing the forward solver to crash.
The numerical advantages of using variational principles are so great, however, that we view this tradeoff as acceptable.
Moreover, there are few physical reasons to prefer one form over another, as long as the asymptotic behavior is correct above and below the critical speed.

>>>>

The most complex and most physically accurate model that we implement is the \emph{first-order} or \emph{Blatter-Pattyn} approximation, a 3D system describing the horizontal velocity \citep{blatter1995velocity, pattyn2003new}.
(In icepack we refer to our implementation of this equation as a \emph{hybrid model} as it includes both shear and plug flow modes, described below.)
The first-order model is based on an asymptotic expansion of the Stokes equations in the ratio of the ice thickness to a typical horizontal length scale.
The aspect ratio of most glacier flows is on the order of 1/20 or less, although there are some exceptions.
For example, the main trunk of Jakobshavn Isbrae in Greenland flows through a very deep and narrow trough with an aspect ratio closer to 1/5.

@@

All the diagnostic models inherit a fundamental nonlinearity in the mechanics of ice flow.
For a Newtonian viscous fluid, the stress tensor $\tau$ and the strain rate tensor $\dot\varepsilon$ are linearly proportional to each other.
Glaciers, however, have a nonlinear constitutive relation:
\begin{equation}
    \dot\varepsilon = A|\tau|^{n - 1}\tau
\end{equation}
where $A$ is the \emph{fluidity} and $n$ is the \emph{Glen flow law exponent} \citep{cuffey2010physics}.
The most commonly-used value of the flow law exponent is $n \approx 3$ as determined from laboratory experiments.
Strictly speaking, the fluidity should be a tensor field, but is almost universally treated as a scalar \citep{gillet2006flow}.

All of the diagnostic models in icepack are described through \emph{variational} or \emph{action} principles \citep{dukowicz2010consistent}.
Rather than describe the velocity as the weak solution of a nonlinear PDE, an action principle instead states that the velocity minimizes a functional called the action.
The action consists of four terms:
\begin{align}
    & \text{action} = \iint\text{stress} \times \text{strain rate}\,dz\,dx - \int\text{basal friction} \times \text{sliding velocity}\,dx \nonumber \\
    & \quad - \iint\text{surface slope}\times\text{velocity}\,dz\,dx - \iint\text{ocean pressure}\times\text{velocity}\,dz\,d\gamma
    \label{action-functional}
\end{align}
where $dz\,dx$ denotes integration over the entire glacier, $dx$ denotes integration over the glacier footprint, and $dz\,d\gamma$ over the side wall boundary.
The action has units of power (energy/time) and can be related to the rate of decrease of the thermodynamic free energy \citep{de2013non}.

@@

The shallow stream model can be derived from the Blatter-Pattyn approximation by assuming nearly plug flow, i.e. $\partial u/\partial z \ll \partial u/\partial x$.
The momentum equations can, again, be vertically integrated to obtain a 2D system of equations.
The class in icepack that represents this physics model is called \texttt{IceStream}.
The terms in the action functional are:
\begin{align}
    \text{viscosity} & = \frac{n}{n + 1}\int_\Omega hA^{-\frac{1}{n}}|\dot\varepsilon(u)|^{\frac{1}{n} + 1}dx \\
    \text{friction} & = \frac{m}{m + 1}\int_\Omega C|u|^{\frac{1}{m} + 1}dx \\
    \text{gravity} & = -\int_\Omega\rho_I gh\nabla s\cdot u\,dx \\
    \text{terminus} & = \frac{1}{2}\int_\Gamma(\rho_I gh^2 - \rho_Wgd^2)u\cdot \nu\,d\gamma
\end{align}
and the action functional is
\begin{equation}
    J = \text{viscosity} + \text{friction} - \text{gravity} - \text{terminus}.
\end{equation}
When the ice is floating, there are two simplifications: the friction coefficient $C$ is 0, and the surface elevation $s$ can be written in terms of the thickness $h$ as
\begin{equation}
    s = (1 - \rho_I / \rho_W)h.
\end{equation}
Additionally, the terminal stress term of the action disappears after applying integration by parts to the gravity term to shift the gradient of the surface elevation over onto the velocity.
Since the action functional for ice shelves has fewer terms than for grounded ice streams, we have defined a separate \texttt{IceShelf} model class.
The ice shelf and ice stream models share common components, i.e. the viscosity and side wall stress.

We verified the correctness of the ice shelf model by checking that numerical results converge at the expected rate to an analytical solution for the velocity with a linearly sloping thickness in a rectangular domain \citep{greve2009dynamics}.
When there is basal friction, the shallow stream equations do not have a simple analytical solution.
To get an exact solution, we chose the ice velocity and thickness a priori and used the computer algebra system SymPy \citep{sympy} to generate a friction coefficient that makes these fields an exact solution.
We then checked that numerical results converge to this manufactured solution at the expected rate \citep{roache2002code}.

@@

Finally, the Gauss-Newton solver defines an approximation to the ``first-order'' part of the objective functional Hessian \citep{pratt1998gauss}.
Each iteration of Gauss-Newton is more expensive than that of BFGS or gradient descent because it requires the solution of a more complex linear system than just the mass matrix.
The Gauss-Newton method converges fastest by far in virtually every test case we have found, in some instances by up to factor of 50.

@@

The original intercomparison project specified that participants could use the Weertman sliding law (equation \eqref{eq:weertman-sliding}) as well as two other sliding laws that transition to a more plastic rheology at high sliding speeds.
The first alternative sliding law consists of Weertman sliding until the stress reaches a critical value, at which point the constitutive relation transitions to perfect plasticity.
The second alternative is the Schoof sliding law \citep{schoof2005effect}:
\begin{equation}
    \tau_b = -\frac{C|u|^{1/m}\cdot \tau_c}{(C^m|u| + \tau_c^m)^{1/m}}\frac{u}{|u|},
    \label{eq:schoof-sliding}
\end{equation}
where the critical stress $\tau_c$ is a certain specified fraction of the water pressure in the subglacial hydrological system.
Sliding laws in icepack are not expressed directly, but rather as the the derivative of an action functional.
To implement the Schoof sliding law we need to know the antiderivative of equation \eqref{eq:schoof-sliding}.
We found the antiderivative of this expression using a computer algebra system, but the result includes hypergeometric functions.
The Unified Form Language has several transcendental functions (sine, cosine, exponential, etc.) but it does not currently support hypergeometric functions.
Instead, we implemented a sliding relation that exhibits the important features of the Schoof law, i.e. it behaves like an $m = 3$ power law at low sliding speeds and $m = \infty$ at high sliding speeds, but which is more tractable algebraically.
Knowing the critical stress $\tau_c$ and the friction coefficient $C$, which has units of stress $\times$ speed${}^{-1/m}$, we can define a \emph{critical speed} $u_c$ as
\begin{equation}
    u_c = C^{-m}\tau_c^m.
\end{equation}
The power dissipation density for the Schoof-type sliding law that we use is
\begin{equation}
    P = \tau_c\left\{\left(u_c^{\frac{1}{m} + 1} + |u|^{\frac{1}{m} + 1}\right)^{\frac{m}{m + 1}} - u_c\right\}
    \label{eq:power-dissipation-modified-schoof-sliding}
\end{equation}
Figure \ref{fig:sliding-laws} shows a comparison of the original Schoof sliding law and the sliding law that arises as the derivative of the functional in equation \eqref{eq:power-dissipation-modified-schoof-sliding}.
The two have the same asymptotic behavior when the speed is much smaller or much larger than the critical speed; they differ in a relatively small range around the critical speed.
The relative difference in basal shear stress between the two sliding laws is less than 10\% throughout the entire range.
Although figure \ref{fig:sliding-laws} shows a comparison of both sliding laws with the same value of the critical speed, by using different values, the agreement between our sliding law can be brought into much closer agreement with the Schoof law.
Finally, the Schoof law itself is phenomenological.
A different sliding law with the right asymptotic behavior is no more or less valid.

@@

Using variational principles to express all constitutive laws is less flexible than specifying the sliding law directly and this is a distinct disadvantage.
We were nonetheless able to implement a sliding law that exhibits the important characteristics of the Schoof sliding law.
\citet{asay2016experimental} also suggest using a sliding law that transitions sharply to exact plasticity above the critical speed.
Expressing this sliding law in UFL requires a conditional in the velocity and is thus no longer differentiable, causing the forward solver to crash.
The numerical advantages of using variational principles are so great, however, that we view this tradeoff as acceptable.
